Towards a Complete Model of Mutation Selection Balance in Humans

### 2023-05-17 manuscript/manuscript.tex
writing

Both the classic background selection and SC16 models imagine a negative
selection process where a continual influx of deleterious mutations flow into
the population at a rate of $\mu$ per basepair per generation in a conserved
region of $L$ basepairs, such that the region-wide per generation mutation rate
is $U = \mu L$. Each mutation imposes an additive selective cost of $s$ in
heterozygotes and $2s$ in homozygotes. The classic background selection model
considers when mutations are sufficiently deleterious that they are destined to
extinction
\parencite{Charlesworth1993-gb,Nordborg1996-nq,Hudson1995-pt,Hudson1994-oh}. In
this case, the genealogy is well-approximated by a neutral coalescence process
with a rescaled effective population size of $N_0 = N \exp(-\nicefrac{U}{s})$.
This approximation works reasonably well for two reasons (though see
\textcite{Cvijovic2018-vd,Walczak2012-fi,Nicolaisen2012-vs}). First, when
selection is strong, the number of deleterious mutations per haplotype has a
stationary Poisson distribution with an average equal to the equilibrium under
the deterministic mutation-selection model \parencite{Haldane1927-ga}. Since
lineages carrying any strongly deleterious mutations cannot contribute to the
genealogy, coalescence events only occur in the fraction
$\exp(-\nicefrac{U}{s}$ of the population free of mutations. Second, the time
it takes for a present-day lineage to trace their ancestry back to a
mutation-free ancestor (the ``delay phase") is negligible compared to the
coalescence timescale of $N_0$ among these ancestors (the ``coalescence
phase"), and can be ignored (\textcite{Durrett2008-ql}, p. 213;
\textcite{Good2014-yz}).

When mutations are weakly deleterious, their allele frequencies are strongly
influenced by stochastic perturbations and their frequency trajectories are no
longer well-approximated by deterministic models. When genetic drift is the
only source of randomness in allele frequency change, the relevant scale of
stochastic perturbations is determined by the drift-effective population size
$N_d$ (e.g. due to only non-selective process). For reasons that will soon be
clear, we consider a general expression of the drift-effective population size
that allows for extra non-heritable variability in offspring number $V^2$,
%
\begin{align}
    \label{eq:simple_Ne}
    N_d = \frac{2N}{V^2 + 2}
\end{align}

where if $V^2 = 2$, this reduces to the classic Wright--Fisher model
\parencite{Wright1938-tv}.

The insight of nearly neutral theory (XXX) was that the qualitative behavior of
selected sites is determined by the compound parameter $2N_d s$. When $2N_d s
\gg 1$, the stochastic component of allele frequency change is minuscule
relative to selection and can be ignored; this is why the factor that rescales
the drift-effective population size under the BGS model does not depend on
$N_d$. However, when $2N_d s \le 1$ stochastic perturbations are on a scale
equal to or greater than the changes due to selection. In this regime, weakly
deleterious alleles can drift up to intermediate frequencies before their
eventual loss or fixation. When this occurs, classic BGS theory breaks down for
several reasons. First, the ``delay phase" is no longer negligible as lineages
carrying weakly deleterious mutations can persist on coalescent timescales.
This distorts genealogies away those expected under neutrality
\parencite{Przeworski1999-mb,OFallon2010-my,Higgs1995-xc}. Second, the
distribution of the number of deleterious mutations (and its corresponding
fitness distribution) are no longer stationary and become a traveling wave
\parencite{Rouzine2008-qz,Good2013-lp,Gessler1995-hz} towards reduced
population fitness (since beneficial mutations are ignored). Consequently,
individuals in the least-loaded class are stochastically lost, fixing those
mutations with a click of ``Muller's ratchet" \parencite{Muller1964-ki}.
Determining both the rate of Muller's ratchet
\parencite{Haigh1978-gt,Gordo2002-dr,Gessler1995-hz} and the shape of
genealogies under weak selection have been stubborn open problems in
evolutionary genetics.

To further complicate matters, both strong and weakly deleterious mutations
create \emph{heritable} fitness variation. In turn, the presence of heritable
fitness variation generates another source of randomness in allele frequency
change known as genetic \emph{draft} \parencite{Neher2013-dz}. Genetic draft
occurs when alleles become randomly associated with different fitness
backgrounds each generation, which stochastically perturbs their trajectories.
Like genetic drift, perturbations due to draft are directionless but overall
increase the variance in allele frequency change. Unlike drift, stochastic
perturbations caused by draft are autocorrelated across generations as long an
allele does not recombine or segregate off its fitness background and the
selective environment does not change
\parencite{Robertson1961-ho,Santiago1995-hx,Buffalo2019-qs}. When selection is
strong, draft can generate multiple-merger coalescences that create
discontinuities in allele frequency trajectories, making diffusion
approximations ill-suited \parencite{Gillespie2000-mh,Der2011-it,Neher2013-dz}.
However, when selection is weak and polygenic, draft can be approximated as an
increase in the rate of drift. The key insight is that the draft rescaling
factor is like extra variability in offspring number, but because of
autocorrelation due to linkage and similar environments, the heritable fitness
variation $C^2$ (on a relative fitness scale, i.e. the squared coefficient of
variance) becomes inflated by a factor $Q^2$. Intuitively, the product $V^2
Q^2$ represents the expected total variance a neutral mutation experiences over
its lifetime in a system with draft. This \emph{heritable} fitness variance
that neutral allele experiences can then be included in Equation
\eqref{eq:simple_Ne} (assuming the usual Wright--Fisher non-heritable variance,
i.e. $V^2=2$),

\begin{align}
    \label{eq:main_Ne}
    N_e = \frac{N}{C^2 Q^2 + 1}.
\end{align}

(c.f. \cite{Robertson1961-ho,Santiago1995-hx}; see XXX for a proof). Note that
because the variance under drift increases each generation due to the
accumulation of drift perturbations and draft autocorrelations, the inflation
term $Q^2$ has a temporal dimension. For notational efficiency, we assume $Q^2$
refers to the asymptotic $Q_t^2$ as $t \to \infty$ and $N_e$ is the
\emph{asymptotic} draft effective population size. In general, however,
statistics such as heterozygosity depend on the cumulative autocorrelation
factor \emph{up to some time} $t$, $Q_t^2$ \parencite[p.
2111]{Santiago1998-bs}. This reflects the fact that effective population size
experienced by a mutation newly associated with a fitness background is larger
than the asymptotic effective population size, since not much autocorrelation
has accumulated in the stochastic perturbations caused by drift yet. From a
backwards-time perspective, this reflects the fact that pairwise diversity
(which is approximately heterozygosity in most cases) depends on the pairwise
coalescence rate per generation, which will is not constant under draft. The
benefit of using a forward-time model of draft is that it the inflation factor
is invariant with respect to the particular fitness background the neutral
allele becomes stochastically associated with. By contrast, the difficulty with
backwards-time models is that the coalescence rates experienced by a lineage
are not invariant to which lineage was sampled.

Equation \eqref{eq:main_Ne}

Shortly, we will show that Robertson (1961) and Santiago and Caballero's (1995,
1998, 2016) approach

rescaling the drift-effective population size by an appropriate factor
\parencite{Robertson1961-ho,Santiago1995-hx,Santiago1998-bs,Barton2000-zg}.
While a simple rescaling cannot elucidate details about how selection alters
the genealogy, it does allow for the calculation of key quantities like the
pairwise coalescence time (though see \textcite{Santiago1998-bs}, p. 2111).

@@

-

@@

or for continuous $t$,

\begin{align}
  \label{eq:q_t}
  Q_t &= 1+\frac{(1-\kappa) (1-r) \left(1-(1-\kappa)^t (1-r)^t\right)}{\kappa+(1-\kappa)r}
\end{align}

which as $t \to \infty$, converges to,

>>>>

Both the classic background selection and SC16 models imagine a negative
selection process where a continual influx of deleterious mutations flow into
the population at a rate of $\mu$ per basepair per generation in a conserved
region of $L$ basepairs, such that the region-wide per generation mutation rate
is $U = \mu L$. Each mutation imposes an additive selective cost of $s$ in
heterozygotes and $2s$ in homozygotes. The classic background selection model
considers when mutations are sufficiently deleterious that they are destined to
extinction
\parencite{Charlesworth1993-gb,Nordborg1996-nq,Hudson1995-pt,Hudson1994-oh}. In
this case, the genealogy is well-approximated by a neutral coalescence process
with a rescaled effective population size of $N_0 = N \exp(-\nicefrac{U}{s})$.
This approximation works reasonably well for two reasons (though see
\cite{Cvijovic2018-vd,Walczak2012-fi,Nicolaisen2012-vs}). First, when selection
is strong, the number of deleterious mutations per haplotype has a stationary
Poisson distribution with an average equal to the equilibrium under the
deterministic mutation-selection model \parencite{Haldane1927-ga}. Since
lineages carrying any strongly deleterious mutations cannot contribute to the
genealogy, coalescence events only occur in the fraction
$\exp(-\nicefrac{U}{s})$ of the population free of mutations. Second, the time
it takes for a present-day lineage to trace their ancestry back to a
mutation-free ancestor (the ``delay phase") is negligible compared to the
coalescence timescale of $N_0$ among these ancestors (the ``coalescence
phase"), and can be ignored (\cite{Durrett2008-ql}, p. 213;
\cite{Good2014-yz}).

When mutations are only weakly deleterious, their allele frequencies are
strongly influenced by stochastic perturbations and their frequency
trajectories are no longer well-approximated by deterministic models. When
genetic drift is the only source of randomness in allele frequency change, the
relevant scale of stochastic perturbations is determined by the drift-effective
population size $N_e$ (e.g. due to only non-selective process). The insight of
nearly neutral theory (XXX) was that the qualitative behavior of selected sites
is determined by the compound parameter $2N_e s$. When $2N_e s \gg 1$, the
stochastic component of allele frequency change is minuscule relative to
selection and can be ignored; this is why the factor that rescales the
drift-effective population size under the BGS model does not depend on $N_e$.
However, when $2N_e s \le 1$, the stochastic perturbations are on a scale equal
to or greater than the changes due to selection. In this regime, weakly
deleterious alleles can drift up to intermediate frequencies before their
eventual loss or fixation. When this occurs, classic BGS theory breaks down for
several reasons. First, the ``delay phase" is no longer negligible as lineages
carrying weakly deleterious mutations can persist on coalescent timescales.
This distorts genealogies away those expected under neutrality
\parencite{Przeworski1999-mb,OFallon2010-my,Higgs1995-xc}. Second, the
distribution of the number of deleterious mutations (and its corresponding
fitness distribution) are no longer stationary and become a traveling wave
\parencite{Rouzine2008-qz,Good2013-lp,Gessler1995-hz} towards reduced
population fitness (since beneficial mutations are ignored). Consequently,
individuals in the least-loaded class are stochastically lost, fixing those
mutations with a click of ``Muller's ratchet" \parencite{Muller1964-ki}.
Determining both the rate of Muller's ratchet
\parencite{Haigh1978-gt,Gordo2002-dr,Gessler1995-hz} and the shape of
genealogies under weak selection have been stubborn open problems in
evolutionary genetics.

To further complicate matters, both strong and weakly deleterious mutations
create \emph{heritable} fitness variation. In turn, the presence of heritable
fitness variation generates another source of randomness in allele frequency
change known as genetic \emph{draft} \parencite{Neher2013-dz}. An allele
experiences genetic draft when it becomes randomly associated with a fitness
background, which stochastically perturbs its trajectory across generations.
Like genetic drift, perturbations due to draft are directionless but overall
increase the variance in allele frequency change. Unlike drift, stochastic
perturbations caused by draft are autocorrelated across generations for as long
an allele does not recombine or segregate off its fitness background and the
selective environment does not change
\parencite{Robertson1961-ho,Santiago1995-hx,Buffalo2019-qs}. When selection is
strong, draft can generate multiple-merger coalescences that create
discontinuities in allele frequency trajectories, making diffusion
approximations ill-suited \parencite{Gillespie2000-mh,Der2011-it,Neher2013-dz}.
However, when selection is weak, draft can be approximated as an increase in
the rate of drift (or equivalently, a reduction in effective population size).

A key insight from Robertson (\citeyear{Robertson1961-ho}) is that in the
long-run draft acts like an increase in the variance in offspring number.
\textcite{Wright1938-tv} showed that additional non-heritable variance in
offspring number (e.g. due to mating strategy) rescales the population size
according to,

\begin{align}
    \label{eq:simple_Ne}
    N_e = \frac{2N}{V + 2}
\end{align}

However, unlike non-heritable variation in offspring number, the heritable
variation is autocorrelated across generations. At the individual level
Robertson considered, this is because offspring from large families tend to
beget many descendents themselves (and likewise with small families). However,
the same process acts at the genomic level, as the perturbations to a neutral
allele's trajectory from its particular fitness background tend to occur in the
same direction across generations. The additional heritable variation in
offspring number is the additive fitness variation $V_A$, which is inflated by
a factor $Q_t^2$ due to the build up of autocorrelation to generation $t$. We
let $Q^2 = Q_t^2$ as $t \to infty$. Intuitively, the product $V_A Q^2$
represents the expected total variance in reproductive success a neutral
mutation experiences over its lifetime in a system with draft. Incorporating
the impact of time-invariant autocorrelation created by draft as rescaled rate
of drift is analogous to using the Green--Kubo relation to find the transport
coefficient \parencite{Green1954-kl,Kubo1957-va}.

Including the total asymptotic variance in reproductive success in Equation
\eqref{eq:simple_Ne} and assuming Wright--Fisher levels of non-heritable
variation (i.e. $V = 2 + 4 Q^2 V_A$),

\begin{align}
    \label{eq:main_Ne}
    N_e = \frac{N}{Q^2 V_A + 1}
\end{align}

(c.f. \cite{Robertson1961-ho,Santiago1995-hx}; see XXX for a proof). Note that
because the variance under drift increases each generation due to the
accumulation of drift perturbations and draft autocorrelations, the inflation
term $Q^2$ has a temporal dimension. For notational efficiency, we let $Q^2$
refer to the asymptotic $Q_t^2$ as $t \to \infty$ and $N_e$ is the
\emph{asymptotic} weak draft effective population size. In general, however,
statistics such as heterozygosity depend on the cumulative autocorrelation
factor \emph{up to some time} $t$, $Q_t^2$ \parencite[p.
2111]{Santiago1998-bs}. This reflects the fact that effective population size
experienced by a mutation newly associated with a fitness background is larger
than the asymptotic effective population size, since not much autocorrelation
has accumulated in the stochastic perturbations caused by drift yet. From a
backwards-time perspective, this reflects the fact that pairwise diversity
(which is approximately heterozygosity when $2N\mu \ll 1$) depends on the
pairwise coalescence rate per generation, which is not constant under draft.
The benefit of using Robertson's forward-time model of draft is that the
inflation factor is invariant with respect to the particular fitness background
the neutral allele becomes stochastically associated with. By contrast, the
difficulty with modeling draft backwards in time is that the coalescence rates
experienced by a lineage are not invariant to which lineage was sampled due to
its particular associated fitness background.

Equation \eqref{eq:main_Ne} is general, since different modes of selection and
linkage can be accommodated by different expressions for the inflation factor
$Q^2$. When fitness variation has a multiplicative polygenic basis, as it does
for genome-wide negative selection processes, the asymptotic weak draft
effective population size experienced by an arbitrary neutral site under the
influence of all linked regions is,

\begin{align}
    \label{eq:polygenic_Ne}
    N_e \approx N \exp\left(-\sum_{i=1}^n V_{A,i} \frac{Q_i^2}{2}\right)
\end{align}

where the factor of one-half comes from ignoring weak associations from
unlinked regions and chromosomes (see XXX). In our genome-wide model, we
consider the summation in Equation \eqref{eq:polygenic_Ne} over non-overlapping
segments $i \in \{1, 2, \ldots, n\}$ each undergoing selection such that
segment $i$ contributes additive fitness variance $V_{A,i}$ to the total
additive genetic fitness variance. The degree to which segment $i$ inflates the
total variance in reproductive success depends on the how quickly the
autocorrelations in

autocorrelation function in
reproductive success. Assuming a constant selective environment, the
autocorrelation in reproductive success of the neutral allele is given by a
geometric autocorrelation function $C(t) = \[(1-r_i)(1-\kappa_i)\]^t$. The
decay in autocorrelation is due to

\begin{align}
    Q_i &= 1 + \sum_{j=1}^\infty \left[(1-r_i)(1-\kappa_i)\right]^j \\
        &= \frac{1}{\kappa_i + r_i(1-\kappa_i)} 
\end{align}

Segment $i$ inflates the total variance in
reproductive success of the neutral allele by $Q_i^2$, according to two key
parameters. First,

according to the chance it stays
associated each generation, given by the recombination fraction $r_i$, and the
extent that the associated heritable fitness variation within the segment
determines future reproductive success in later generations. Assuming a
constant environment for selection, this

, according to
the recombination fraction $r_i$ to the segment and the extent to which the
fitness variation in the segment perturbs the trajectory in later generations.

the fitness
variation associated with the neutral allele

persistence of the
associated fitness variance through the generations.

Here, we consider each $i$ to refer
to an $L_i$-basepair segment under negative selection (though this model
accommodates other forms of polygenic selection, \cite[p.
2109]{Santiago1998-bs}). The inflation factor $Q_i^2$ depends on the
recombination fraction $r_i$ between the neutral site and selected segment, and
the rate at which the fitness variance at segment $i$ associated with the

neutral allele decays across generations, $1-\kappa_i$. This associated fitness
variation decays at rate $1-\kappa_i$ due only to the dynamics of polygenic
selection within the segment \parencite{Bulmer1980-zo,Turelli1990-kd}, assuming
a constant environment. Then, the root of the asymptotic variance inflation
factor $Q_i$ experienced by the neutral allele is,

\begin{align}
    Q_i &= 1 + \sum_{j=1}^\infty \left[(1-r_i)(1-\kappa_i)\right]^j \\
        &= \frac{1}{\kappa_i + r_i(1-\kappa_i)} 
\end{align}

(see Appendix Equation \ref{eq:Qinf}). This general equation can accommodate
models of polygenic selection as long as the equilibrium additive fitness
variation $C_i^2$ can be specified and the change in variance due to selection
can be approximated as a geometric decay, i.e. $\Delta C_i^2 = -\kappa C_i^2$
\parencite{Bulmer1971-ae,Keightley1988-eq,Walsh2018-bt}. This is usually a
reasonable assumption since within-generation selection removes a fraction of
phenotypic variation from the population, and the fraction of that phenotypic
variation that is genetic is approximately the heritability $h^2$.

Assuming the fitness
variation $C_i^2$ in the segment is at an equilibrium between mutation and
selection, $\Delta C_i^2 = 0$, and $\Delta C_i^2 = V_{m,i} - \kappa C_i^2$

Note that
the decay in fitness variation due to the substitution of weakly deleterious
mutations happens on a slower timescale and is ignored here, but as we will
impacts the overall level of $C_i^2$.

@@

- we need to talk about Fisher's Fundamental Theorem and how this model
connects fitness variation to absolute changes in fitness.

@@

% or for continuous $t$,

% \begin{align}
%   \label{eq:q_t}
%   Q_t &= 1+\frac{(1-\kappa) (1-r) \left(1-(1-\kappa)^t (1-r)^t\right)}{\kappa+(1-\kappa)r}
% \end{align}
which as $t \to \infty$, converges to,
